Long name of the paper

### 2018-10-05 anzjs2018/sections/03_models.tex
Likeilihood update

After mutating the particle set, the posterior distribution is obtained by
\emph{selection}, which is implemented by importance resampling using
likelihood weights.
To do so, we need to define the measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection,
which allows us to place a bivate normal likelihood on the data.
The model for the observation generation is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}

% The vehicle model involves infering the \emph{unobservable state} $\bX_k$ 
% of a vehicle at time $t_k$ from the \emph{observed state} $\by_k$,
% which in this case is the GPS location of the vehicle,
% where $\lambda_k$ and $\phi_k$ are the longitude and latitude, respectively.
% \begin{equation}
% \label{eq:vehicle_observation}
% \by_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}
% \end{equation}
% To construct the unobservable state, 
% each vehicle is assumed to follow the trajectory defined by a trajectory function of time, $x(t)$,
% such that the distance traveled along a route by a vehicle at time $t_k$ is given by
% $x_k = x(t_k)$.
% Additionally, speed and acceleration are given by
% the first and second derivatives of the trajectory function,
% $\dot x_k = x'(t_k)$ and $\ddot x_k = x''(t_k)$, respectively.
% Exactly how the state is constructed will depend on the model being used.

% Since observations of the vehicle are irregular, 
% a \emph{transition function} $f$ is used to
% predict the vehicle's state at time $t_k$ given the current knowledge of the state at time $t_{k-1}$,
% as well as accounting for \emph{system noise}, $\sigma_x^2$.
% We also define a \emph{measurement function} $h$ which describes the relationship between
% the observable and unobservable states,
% and accounting for \emph{measurement error}, $\bv_k$,
% \begin{equation}
% \label{eq:vehicle_measurement}
% \by_k = h(\bX_k) + \bv_k = h(x_k) + \bv_k,
% \end{equation}
% so all models need to estimate at least $x_k$, the distance traveled along the route.

% The model is split into two steps, prediction and update.
% The prediction involves using the transition function to predict
% the next state, $\hat\bX_{k|k-1}$,
% while the update step using a likelihood function and the measurement function
% to update the state to account for the observed vehicle location, $\by_k$,
% giving a final estimate of $\hat\bX_{k|k}$.

% \subsubsection{Predicting state using the transtion function}
% \label{sec:pf_prediction}

% The transition function is a model of bus behaviour that allows
% predictions of future states to be made given the most recently 
% updated state.
% Developing a good model for bus behaviour allows us to estimate
% parameters of interest, such as vehicle speed,
% dwell time at bus stops, and travel time along individual roads,
% the last of which is of course our primary objective.
% In this section, we propose [HOW MANY?] increasingly complex models
% of bus behaviour.

% %% MODEL 1
% The first model, $\bM_0$, assumes a constant speed between observations
% using the state
% $\bX_k = \left[\begin{smallmatrix}x_k && \dot x_k\end{smallmatrix}\right]^\top$
% and Gaussian system noise with mean zero and variance adjusted by
% $\Delta_k = t_k - t_{k-1}$.
% \begin{equation*}
% \hat\bX_{k|k-1} = f_0(\bX_{k-1|k-1}, \Delta_k, \sigma_x^2) =
% \begin{bmatrix}
% x_{k|k-1} & \dot x_{k|k-1}
% \end{bmatrix}
% \end{equation*}
% where
% \begin{align*}
% x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k|k-1} \\
% x_{k|k-1} &= \dot x_{k-1|k-1} + w_k \\
% w_k &\sim \mathcal{N}_T(0, \Delta_k \sigma_x^2)
% \end{align*}
% Here, $\mathcal{N}_T$ denotes a truncated normal distribution such that
% the system noise is truncated to ensure the speed remains between 0 and 30~m/s.

% While $\bM_0$ may perform well on average,
% and indeed if all we were interested in was estimating $x_k$ it may be adequate;
% however, it is missing two important features of bus behaviour:
% \begin{itemize}
% \item variable speeds along roads, \emph{which is what we are truly interested in}
% \item unknown dwell times at bus stops
% \end{itemize}

% The first of these is incorporated into $\bM_1$,
% which allows vehicle speeds to vary over time.
% This model uses the same state as $\bM_0$,
% but a different transition function $f_1$.
% In this case, the state equations depend on a sequence of system noise,
% $\bw_k = \{\bw_k^1, \cdots, \bw_k^{\Delta_k}\}$
% \begin{align}
% \label{eq:transition_f1}
% x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \nonumber \\
% \dot x_{k|k-1} &= \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
% w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2) \nonumber
% \end{align}
% Again, system noise is assumed to have a Gaussian distribution, 
% truncated to ensure that the vehicle's speed is always between 0 and 30~m/s.
% This is achieved by iteratively estimating each second the the vehicle's state,
% each time ensuring the speed remains in the specified range.

% The next model, $\bM_2$, allows for dwell times at bus stops.
% Since buses can either stop and wait, or not stop,
% multimodality is introduced to the system,
% which requires a modeling framework, such as the particle filter used here,
% to handle it.
% For each bus stop $j = 1, \ldots, M$ along a route,
% its distance into the trip, $S_j^d$, 
% is used to determine when a vehicle passes a stop.
% When this occurs, the dwell time $\delta_j \geq 0$ at stop $j$ is modelled
% using the stopping probability, $\pi_j$,
% a common dwell time parameter $\gamma$ which models the
% deceleration, opening and closing of doors, and acceleration at bus stops,
% and a stop dwell time $\tau_j$ for passengers alighting and disembarking.
% Therefore the stop dwell time is
% \begin{equation}
% \label{eq:dwell_time}
% \delta_j = p_j(\gamma + \tilde\delta_j),\quad
% p_j \sim \mathrm{Bern}(\pi_j),\quad
% \tilde\delta_j \sim \mathcal{E}(\tau_j)
% \end{equation}

% To define the transition function for this model, $f_2$,
% we need to account for multiple dwell times.
% Let $s_n$ be the index of the next stop,
% and sample dwell times for each upcoming stop $j = s_n, \ldots, M$,
% noting that these can be zero.
% Assuming the vehicle passes stops $s_n$ up to $J-1$,
% does it reach stop $J$ in the time remaining?
% \begin{equation}
% \tilde X_k^J = \min\left\{
% S_J^d, f_1(\hat X_{k|k-1}, \Delta_k - \sum_{j=s_n}^{J-1}\delta_j)\right\}
% \end{equation}
% Taking the maximum of the sequence gives the predicted distance
% traveled by the vehicle, so
% \begin{equation}
% \hat X_{k|k-1} = f_2 (\hat X_{k-1|k-1}, \Delta_k) =
% \max_{J = s_n, \ldots, M}\left\{
%     \tilde X_k^J
% \right\}
% \end{equation}

% Care needs to be made when generating the sequence,
% so in practice the state will be calculated recursively to ensure
% the system noise is consistent.
% \begin{equation*}
% \hat{\tilde X}_k^J =
% \begin{cases}
%     f_1(\hat{\tilde X}_k^{J+1}, \delta_J) & J < M \\
%     \hat X_{k-1|k-1} & J = M
% \end{cases}
% \end{equation*}

% The final model $\bM_3$ demonstrated here will allow acceleration to vary.
% \begin{align*}
% \label{eq:model_m3}
% x_{k|k-1} &= x(t_k) \\
% \intertext{where states are recursively calculated back to $t_{k-1}$}
% x(t) &= x(t - 1) + \dot x(t) \\
% \dot x(t) &= \dot x(t-1) + \ddot x(t) \\
% \ddot x(t) &= \ddot x(t-1) + w_t,\quad
% w_t \sim \mathcal{N}(0,\sigma_x^2)
% \end{align*}
% By truncating the acceleration, 
% the speed of the vehicle can be maintained between 0 and 30~m/s.
% Additionally, when approaching a bus stop that will be stopped at,
% the acceleration will be trunctated to below zero,
% however no physical deceleration will be modeled as this is too complex
% and stopping behaviour of buses can be sudden 
% (for example when passengers signal late that they wish to board or disembark).
% After servicing a stop, the bus will need to accelerate up to speed again.

% \subsubsection{Updating state using the observation likelihood}
% \label{sec:pf_update}

% The update step involves comparing the predicted state $\bX_{k|k-1}$
% with the observed state $\bY_k$,
% using the measurement function $h$ to convert between the two state spaces,
% and a likelihood function.
% The measurement function is a deterministic function that uses only the 
% distance traveled, $x_k$, along a known path $\bR$ to obtain a GPS coordinate.
% The inverse measurement function, which is used in other approaches,
% is \emph{not deterministic}, and instead uses an optimisation to
% find the ``estimated'' distance traveled,
% which invariably leads to issues in the implementation.
% By using a particle filter approach, we avoid using $h^{-1}$
% and instead determine the liklihood using $\bY_k$ directly.

% For the likelihood, the GPS error is assumed symmetrical about the true
% position of the vehicle, with a radius of $\sigma_y^2$ meters.
% Define a projection function $g(\cdot | \bY)$ that projects points
% onto a cartesian surface (``Flat Earth''),
% so that the distance between points is the real Earth distance.
% Then the distance between the points is
% \begin{align*}
% \delta_k &= \sqrt{\left(g(h(X_k))_1 - g(\bY_k)_1\right)^2 +
%     \left(g(h(X_k))_2 - g(\bY_k)_2\right)^2} \\
% \intertext{which can be standardised by the GPS error}
% \tilde \delta_k &=
% \sqrt{\left(\frac{g(h(X_k))_1 - g(\bY_k)_1}{\sigma_y}\right)^2 +
%     \left(\frac{g(h(X_k))_2 - g(\bY_k)_2}{\sigma_y}\right)^2} =\frac{\delta_k}{\sigma_y} \\
% \intertext{Since we assume symmetric variance, this is actually}
% \tilde\delta_k &= \sqrt{Z_1^2 + Z_2^2}, \quad
% Z_1,Z_2 \sim \mathcal{N}(0,1)
% \end{align*}
% The sum of two normal random variables has an Exponential distribution with rate $\frac{1}{2}$,
% therefore 
% \begin{equation}
% \label{eq:lhood_exp}
% \left(\frac{\delta_k}{\sigma_y}\right)^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
% \end{equation}

% Combining this together, we use a distance function $d(\bY_1, \bY_2)$ that calculates
% the geographic distance between two GPS coordinates,
% the likelihood is
% \begin{equation}
% \label{eq:lhood}
% f(\bY_k | \hat\bX_{k|k-1}, \sigma_y^2) =
% \frac{1}{2} \exp\left\{-\frac{d(\bY_k, h(\hat\bX_{k|k-1}))^2}{2\sigma^2}\right\}
% \end{equation}

>>>>

After mutating the particle set, the posterior distribution is obtained by
\emph{selection}, which is implemented by importance resampling using
likelihood weights.
To do so, we need to define the measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection,
which allows us to place a bivate normal likelihood on the data.
The model for the observation generation is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
However, due to the assumption that error in longitude and latitude are uncorrelated, 
the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent standard normal random variables $z_1$ and $z_2$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}
\end{equation}
As the sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is therefore easy to calculate using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible within the particle filter;
in other approaches, such as the Kalman filter,
the likelihood cannot be calculated using the distance as the estimate
is a distribution, and so instead a reverse non-deterministic transformation
is required, which introduces additional error and uncertainty into the model.

The posterior distribution is obtained by \emph{selection},
or resampling with replacement using likelihood weights
\begin{equation*}
W_k^{(i)} = \frac{p(\bY_k | \bX_k^{(i)}, \epsilon)}{
    \sum_{j=1}^N p(\bY_k | \bX_k^{(j)}, \epsilon)
} 
W_{k-1}^{(i)} 
\end{equation*}
However, to improve efficiency, the state $\tilde \bX_k$ is only resampled
when the effective sample size 
$N_{\text{eff}}$ is less than some threshold,
\begin{equation*}
N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (W_k^{(i)})} < N_{\text{thres}}
\end{equation*}
When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.
